---
title: "Customer Churn Analysis"
author: "Shahbaz"
date: "2025-12-18"
output: html_document
---

## Summary

This project analyzes customer churn behavior using R, with the objective of identifying key churn drivers and building predictive models to flag high-risk customers. Since an explicit churn label was not available in the dataset, churn was defined using domain-driven behavioral rules based on customer tenure, usage frequency, support interactions, and payment delays.

The analysis begins with exploratory data analysis (EDA) to understand customer behavior and churn patterns across subscription types and contract lengths. Feature engineering was performed to construct a binary churn indicator, followed by the calculation of churn rate as a key business metric.

Predictive modeling was conducted using logistic regression and decision tree classifiers. A train-test split was applied to ensure proper model evaluation. Model performance was assessed using confusion matrix metrics and ROC-AUC, demonstrating strong predictive capability. The project concludes with churn driver analysis and identification of high-risk customers to support targeted retention strategies.


# Loaded Necessary Libraries 
```{r}
library(tidyverse)
library(ggplot2)
library(caret)
library(rpart)
library(rpart.plot)
library(knitr)
library(dplyr)
```

# Data Loading
```{r}
churn <- read.csv("/Users/shahbazshaikh/Downloads/customer_churn_dataset-testing-master.csv")
head(churn)
```

# Summarization
```{r}
str(churn)
summary(churn)
```

# Feature Engineering – Create Churn Label
```{r}
churn <- churn %>%
  mutate(
    Churn = ifelse(
      (Tenure < 12 & Usage.Frequency < 5) |
      (Support.Calls >= 4) |
      (Payment.Delay > 20),
      1, 0
    )
  )

churn$Churn <- factor(churn$Churn, labels = c("No", "Yes"))
```

# Churn Rate KPI
```{r}
churn_rate <- mean(churn$Churn == "Yes") * 100
churn_rate
```

# Churn by Contract Length
```{r}
ggplot(churn, aes(Contract.Length, fill = Churn)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Churn Rate by Contract Length")
```

# Churn by Subscription Type
```{r}
ggplot(churn, aes(Subscription.Type, fill = Churn)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Churn Rate by Subscription Type")
```


# Train-Test Split
```{r}
set.seed(123)

train_index <- createDataPartition(churn$Churn, p = 0.7, list = FALSE)
train_data <- churn[train_index, ]
test_data  <- churn[-train_index, ]
```

# Logistic Regression Model
```{r}
log_model <- glm(
  Churn ~ Tenure + Usage.Frequency + Support.Calls +
    Payment.Delay + Subscription.Type + Contract.Length,
  data = train_data,
  family = "binomial"
)

summary(log_model)
```

# Predicting Churn on Test Data
```{r}
test_prob <- predict(log_model, test_data, type = "response")

test_pred <- ifelse(test_prob > 0.5, "Yes", "No")
test_pred <- factor(test_pred, levels = c("No", "Yes"))
```

# Confusion Matrix Evaluation
```{r}
confusionMatrix(test_pred, test_data$Churn, positive = "Yes")
```


# ROC Curve & AUC
```{r}
library(pROC)
roc_obj <- roc(
  response = test_data$Churn,
  predictor = test_prob,
  levels = c("No", "Yes")
)

plot(roc_obj, main = "ROC Curve for Churn Prediction")
auc(roc_obj)
```

# Decision Tree Model
```{r}
tree_model <- rpart(
  Churn ~ Tenure + Usage.Frequency + Support.Calls +
    Payment.Delay + Subscription.Type + Contract.Length,
  data = churn,
  method = "class"
)

rpart.plot(tree_model)
```

# Customer Engagement Visualization
```{r}
ggplot(churn, aes(Tenure, Usage.Frequency, color = Churn)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Usage Frequency vs Tenure (Colored by Churn)")
```

# Churn Driver Analysis
```{r}
churn %>%
  filter(Churn == "Yes") %>%
  summarise(
    Low_Usage = sum(Usage.Frequency < 5),
    High_Support = sum(Support.Calls >= 4),
    Payment_Issues = sum(Payment.Delay > 20)
  ) %>%
  pivot_longer(everything()) %>%
  ggplot(aes(name, value, fill = name)) +
  geom_col() +
  labs(title = "Top Reasons for Churn")
```

# High-Risk Customer Identification
```{r}
high_risk <- churn %>%
  filter(
    Tenure < 12,
    Usage.Frequency < 5,
    Support.Calls >= 3
  )

kable(head(high_risk))
```

## Conclusion

The churn prediction model achieved strong performance, with high overall accuracy and excellent recall for churned customers. The logistic regression model successfully identified the majority of at-risk customers while maintaining high precision, making it suitable for real-world churn prevention use cases where minimizing missed churners is critical.

Key factors contributing to churn include low usage frequency, short customer tenure, frequent support interactions, and payment delays. These insights suggest that early-stage engagement, proactive customer support, and improved billing processes could significantly reduce churn rates.

Although the churn label was rule-based rather than observed, this project demonstrates a complete end-to-end analytics workflow, including data preparation, exploratory analysis, predictive modeling, evaluation, and actionable insight generation. With real churn outcomes, the same framework could be extended into a production-ready churn prediction system.